Realizing temporal graphs from fastest travel times

### 2023-05-05 TGR Extended - New paper/Realizing TG - extended.tex
Poly-time algorithm for paths

\usepackage{amsmath}
%\let\proof\relax
%\let\endproof\relax

@@

\begin{proof}[Proof idea for \cref{thm:paths-upperBound} -- new:]
We are given an instance  $(G,D)$ of \deltaUpperBound\, where $G$ is a path on vertices $v_0, v_1, \dots, v_n$. 
Let $\lambda$ be a labeling satisfying this instance.
We denote with $\tau_i^+$ the waiting time at vertex $v_i$ when passing from vertex $v_{i-1}$ to $v_{i+1}$ in $(G,\lambda)$, with $\tau_i^-$ the waiting time at $v_i$ when traversing through $v_i$ from the other direction (for details see~\cref{def:travelDelays}).
Since $\tau_i^- = \Delta - \tau_i^+$, we denote with $\tau_i = \tau_i^+$, and replace $\tau_i^-$ with $ \Delta - \tau_i$.
By the definition it also follows that the duration of a fastest path among two vertices $v_i, v_j$ in $(G,\lambda)$, for $i < j$
is 
\begin{equation}\label{eq:path-positive}
    d(v_i,v_j)=\sum_{k = i+1}^{j-1} \tau_k + 1,
\end{equation}
and for $i > j$ is 
\begin{equation}\label{eq:path-negative}
d(v_i,v_j)=\sum_{k = i+1}^{j-1} (\Delta - \tau_k) + 1.
\end{equation}

We want to determine the values of all $\tau_i$, and consequently the labeling $\lambda$, such that the distance matrix $D$ is satisfied,
namely $d(v_i,v_j) \leq D_{i,j}$ for all $v_i,v_j$ in $G$.
To do this we build the following integer linear program (ILP).

For every pair of vertices $v_i, v_j \in G$ where $|i-j| \geq 2$, we create an inequality of form $d(v_i,v_j) \leq D_{i.j}$, where $d(v_i,v_j)$ equals the expression from \cref{eq:path-positive} if $i < j$, and \cref{eq:path-negative} if $i > j$.
We build a system of $O(n^2)$ linear inequalities with $n$ variables $d_1,d_2, \dots, d_n$,
where each $d_i \in [\Delta]$. Let us denote this system of inequalities as $I_D$.
The produced system $I_D$ is an ILP. 
We need to show that we can, in polynomial-time, determine if $I_D$ admits a solution,
and in the positive case construct one such solution.
A solution for $I_D$ determines the values of each $\tau_i$. We then use this information to determine a labeling $\lambda$, which is a solution for our instance of \deltaUpperBound.
We do this using the following steps.
First we prove that our ILP $I_D$ has a solution if and only if the (relaxed) LP version of it admits one .
Second, we prove that if we have a (non-necessary integer) solution $S'$ of the LP instance $I_D$, then we can modify the solution $S'$ to get an integer solution $S$ of the ILP version of $I_D$.
\begin{claim}
    Integer version of the LP for $I_D$ has a solution if and only if the LP for $I_D$ has a solution.
\end{claim}

\begin{claimproof}
Clearly, if the LP for $I_D$ admits no solution, then also the Integer version of the LP for $I_D$ admits no solution, as any integer solution for ILP would also be a solution for LP.
Therefore we only need to prove that if the LP for $I_D$ admits a solution, then also the ILP or $I_D$ admits one (not necessarily the same one).

\end{claimproof}

\begin{proof}[Proof idea for \cref{thm:paths-upperBound} -- old:]
	Let $P = (v_1, v_2, \dots, v_n)$ be a path on $n$ vertices.
	For the sake of simplicity we denote with $a_i$ label of edge $e_i =v_iv_{i+1}$, for all $i \in [n-1]$.
	By the definition, for any two vertices $v_i,v_j \in P$, we have the following inequality:
	\begin{equation*}
		(a_{i+1} - a_i)_\Delta + (a_{i+2} - a_{i+1})_\Delta + \cdots + (a_{j-1} - a_{j-2})_\Delta + 1
		\leq D_{i,j}.
	\end{equation*}
	For the sake of simplicity we refer to the inequality that a temporal path from $v_i$ to $v_j$ has to respect (\ie~the above inequality) 
	as $I_{i,j}$.
	Note that we can rewrite the inequality $I_{i,j}$ also without the modulo $\Delta$ part as:
	\begin{equation*}
		k_{i+1,i} \Delta + (a_{i+1} - a_i)+ k_{i+2,i+1} \Delta + (a_{i+2} - a_{i+1}) + \cdots + k_{j-1,j-2} \Delta + (a_{j-1} - a_{j-2})+ 1
		\leq D_{i,j},
	\end{equation*}
	where all $k_{i,i+1} \in \{0,1\}$.

@@

>>>>

\usepackage{amsmath}
\usepackage{cite}
%\let\proof\relax
%\let\endproof\relax

@@

\begin{proof}
We are given an instance  $(G,D)$ of \deltaUpperBound\, where $G$ is a path on vertices $v_0, v_1, \dots, v_{n+1}$. 
Let $\lambda$ be a labeling satisfying this instance.
We denote with $\tau_i^+$ the waiting time at vertex $v_i$ when passing from vertex $v_{i-1}$ to $v_{i+1}$ in $(G,\lambda)$,
and with $\tau_i^-$ the waiting time at $v_i$ when traversing through $v_i$ from the other direction (for details see~\cref{def:travelDelays}).
Since $\tau_i^- = \Delta - \tau_i^+$, we can simplify the above notation and replace $\tau_i^+$ with $\tau_i$, and $\tau_i^-$ with $ \Delta - \tau_i$.
Therefore, by the definition, it follows that the duration of a fastest path among two vertices $v_i, v_j$ in $(G,\lambda)$, for $i < j$
is 
\begin{equation}\label{eq:path-positive}
    d(v_i,v_j)=\sum_{k = i+1}^{j-1} \tau_k + 1,
\end{equation}
and for $i > j$ is 
\begin{equation}\label{eq:path-negative}
d(v_i,v_j)=\sum_{k = i+1}^{j-1} (\Delta - \tau_k) + 1.
\end{equation}

We want to determine the values of all $\tau_i$, and consequently the labeling $\lambda$, such that the distance matrix $D$ is satisfied,
namely $d(v_i,v_j) \leq D_{i,j}$ for all $v_i,v_j$ in $G$.
To do this we build a system of inequalities $I_D$, using the following procedure.

For every pair of vertices $v_i, v_j \in G$ where $|i-j| \geq 2$, we create an inequality of form $d(v_i,v_j) \leq D_{i.j}$, where $d(v_i,v_j)$ equals the expression from \cref{eq:path-positive} if $i < j$, and \cref{eq:path-negative} if $i > j$.
This results in creating $O(n^2)$ linear inequalities with $n$ variables $\tau_1,\tau_2, \dots, \tau_n$,
where each $\tau_i \in [\Delta]$.

Let us observe the following properties of the system of inequalities $I_D$, which will help us to express it as a linear program (LP) in a canonical form.

From observing vertices $v_i$ and $v_j$, we get the following two inequalities:
\begin{equation}
\sum_{k = i+1}^{j-1} \tau_k + 1 \leq D_{i,j}
\Leftrightarrow
\sum_{k = i+1}^{j-1} \tau_k \leq D_{i,j} - 1
\end{equation}
and
\begin{equation}
\sum_{k = i+1}^{j-1} (\Delta - \tau_k) + 1
\Leftrightarrow
\sum_{k = i+1}^{j-1} - \tau_k \leq  D_{i,j} - 1 -(j-i+1)\Delta .
\end{equation}
From the condition $\tau_i \in [\Delta]$, we add the inequality $\tau_i \leq \Delta$ and $\tau_i \geq 0$.

Let us now describe our LP.
We want to find an assignment of values for each $\tau_i$, that satisfy the inequalities:
\begin{equation} \label{eq:paths-inequalities-A}
\begin{aligned}
\tau_1 & \leq \Delta \\
\tau_2 & \leq \Delta \\
& \vdots \\
\tau_n &\leq \Delta \\
%%%% \tau \leq Delta
\tau_1 &\leq D_{0,2} -1 \\
\tau_1 + \tau_2 &\leq D_{0,3} -1 \\
\tau_1 + \tau_2 + \tau_3 &\leq D_{0,4} -1 \\
& \vdots \\
\tau_1 + \cdots + \tau_n &\leq D_{0,n+1} -1 \\
%%%% all with \tau_1
\tau_2 & \leq D_{1,3} -1 \\
\tau_2 + \tau_3 &\leq D_{1,4} - 1\\
& \vdots \\
\tau_n &\leq D_{n-1,n+1} - 1 \\
%%% all positive tau
%%%
- \tau_1 &\leq D_{2,0} -1 - \Delta \\
- \tau_1 - \tau_2 &\leq D_{3,0} -1 -2 \Delta \\
& \vdots \\
- \tau_n &\leq D_{n+1,n-1} - 1 - \Delta
\end{aligned}
\end{equation}
where all $\tau_i$ are positive integers.
We can write the above set of inequalities in the matrix form as $A t \leq b$, where $t$ is a vector of variables $\tau_1,\tau_2,\dots, \tau_n$.
Therefore we can form the following LP, in the canonical form
\begin{equation} \label{eq:paths-LP}
\begin{aligned}
\text{find a vector} &&& t \\
\text{that maximizes} &&& \sum_i \tau_i \\
\text{subject to} && A t  &\leq b, \\
\text{and} && t &\geq 0. \\
\end{aligned}
\end{equation}
Note that in our case we do not necessarily want to maximize the sum of all $\tau_i$ variables, but just to find an assignment of variables that will satisfy our conditions. Clearly if there is a solution to our system then there is also a solution that will maximize the sum, and vice versa.

In their work Hoffman and Kruskal~\cite{Hoffman1956Integral} prove that if the coefficient matrix  $A$, from the \cref{eq:paths-LP} is totally unimodular (that is, any square submatrix has determinant $\pm 1$, or $0$), and the vector $b$ is integral,
then the extreme (corner) points of the feasible polyhedron $\{ t \mid A t  \leq b, t \geq 0\}$ are integral. 
Which implies that using the simplex method to solve the LP, which takes polynomial time, results in an integral solution.
Therefore we have to prove that $A$ really is totally unimodular.

\begin{claim}
    The coefficient matrix $A$ of linear program $I_D$ is totally unimodular.
\end{claim}

\begin{claimproof}
    From the definition it follows that $U$ is a totally unimodular matrix if and only if $U^T$ is totally unimodular.
    It is not hard to prove that if $U$ is a totally unimodular matrix then also $[I U]$ and $[U -U]$ are totally unimodular, where $I$ is the identitiy matrix.
    We use these facts to prove that our coefficient matrix $A$ is totally unimodular.

    From the first $n$ inequalities from \cref{eq:paths-inequalities-A} it follows that the first $n$ rows of $A$ form an identity matrix $I$ of size $n \times n$.
    Let us now split the rest of the matrix $A$ into two parts, one where all the non-zero coefficient are positive ($+1$) and the second where they are negative ($-1$).
    Note that the second part (with the negative non-zero coefficients) is the same as the first part, where each coefficient changes the sign.
    Therefore we can write the matrix $A$ as
    \begin{equation*}
    A =
        \begin{bmatrix}
        I \\
        B \\
        -B
        \end{bmatrix},
    \end{equation*}
    where $I$ is the identity matrix and $B$ is a matrix of the following form
    \begin{equation*}
    B =
        \begin{bmatrix}
        1 & & & & \\
        1 & 1 & & \\
        1 & 1 & 1 & \cdots & 1 \\
        \vdots & & & \ddots & \\
        1 & 1 & 1 & \cdots & 1\\
        & 1 & & & \\
        & 1 & 1 & & \\
        & \vdots & & \ddots & \\
        & 1 & 1 & \cdots & 1\\
        & & & \ddots & \\
        & & & & 1
        \end{bmatrix}.
    \end{equation*}
    Note that $B$ is a matrix in which for every row all non-zero entries ($1$-s) appear consecutively.
    It is known\cite{Hoffman1956Integral,Fulkerson1965Incidence} that $B$ is a totally unimodular matrix.
    Using this result and the above mentioned properties of totally unimodular matrices, it follows that $A$ is totally unimodular.
\end{claimproof}
We have now proved that we can, in polynomial time, determine the values of $\tau_i$ satisfying inequalities from $I_D$, if such $\tau_i$ exist.
We now have to determine the labeling $\lambda$ from the waiting times $\tau_i$.
Note that knowing the label $\lambda(v_{i-1}v_i)$ and $\tau_i$ we uniquely determine the label $\lambda(v_iv_{i+1})$ as $(\tau_i - \lambda(v_{i-1}v_i))\pmod{\Delta}$.
Therefore the algorithm to build the whole labelling $\lambda$ of $G$
starts by fixing the label of the first edge $\lambda(v_0v_1) = 1$, and then iteratively determines the label of each consecutive edge.
At the end we have build a labeling $\lambda$ that satisfies all inequalities from $I_D$, which implies that for all $v_i,v_j \in V(G)$ we have $d(v_i,v_j) \leq D_{i,j}$.
\end{proof}

@@

\clearpage
\section{Old (not useful?) results}
\begin{proof}[Proof idea for \cref{thm:paths-upperBound} -- old:]
	Let $P = (v_1, v_2, \dots, v_n)$ be a path on $n$ vertices.
	For the sake of simplicity we denote with $a_i$ label of edge $e_i =v_iv_{i+1}$, for all $i \in [n-1]$.
	By the definition, for any two vertices $v_i,v_j \in P$, we have the following inequality:
	\begin{equation*}
		(a_{i+1} - a_i)_\Delta + (a_{i+2} - a_{i+1})_\Delta + \cdots + (a_{j-1} - a_{j-2})_\Delta + 1
		\leq D_{i,j}.
	\end{equation*}
	For the sake of simplicity we refer to the inequality that a temporal path from $v_i$ to $v_j$ has to respect (\ie~the above inequality) 
	as $I_{i,j}$.
	Note that we can rewrite the inequality $I_{i,j}$ also without the modulo $\Delta$ part as:
	\begin{equation*}
		k_{i+1,i} \Delta + (a_{i+1} - a_i)+ k_{i+2,i+1} \Delta + (a_{i+2} - a_{i+1}) + \cdots + k_{j-1,j-2} \Delta + (a_{j-1} - a_{j-2})+ 1
		\leq D_{i,j},
	\end{equation*}
	where all $k_{i,i+1} \in \{0,1\}$.
